---
title: "Exam 02: Takehome Portion (Solutions)"
output:
  html_document:
    theme: cosmo
    highlight: zenburn
    css: "../css/note-style.css"
---

## Instructions

Deadline: Monday, 25 March 2024 (Start of Class)

This exam is the same format as the notebooks we have been completing in class.
In order to complete it correctly, please keep in mind that:

1. In each section there is exactly one data set that you need to create. 
You should not give this a name; instead just make sure the table prints out
in the knit file.

2. There are nine questions on this exam. Each is worth 10 points. An 
additional 5 points is assigned based on your code formatting across the
entire exam. 

3. You must submit two files: the Rmarkdown file and a knit version of the
file as an HTML document. These must be uploaded to Box by the deadline. Note
that these will not preview correctly on Box, so be careful when uploading.

4. You may use any static resources, such as course notes and external
websites, but you may not discuss the exam with classmates anyone else.

5. I am happy to answer clarifying questions about the exam or to help you
with unexpected R errors. However, I will not answer content-based questions 
after the exam is posted. Note that I may not be able to answer questions 
sent after noon the day the exam is due.

6. The exam should take no more than 3 hours, but you may use as much time 
as you need.

7. Personal computer issues is not an excuse for not completing the exam on
time. As a back up, you can use the computers in the Jepson computer lab or
in the library. 

Good luck!

## Tropical Storms

The data for this exam consists of information about tropical storms in the 
Atlantic Ocean between 1950 and 2020. The main data set is here; it gives one
row each time a particular storm is measured (usually every 6 hours). Note that
like the movies data, you will need to use both the year and name of the storm
to uniquely describe it.

```{r}
source("../funs/funs.R")
storms <- read_csv("../data/storms.csv")
storms
```

The variables are:

- **year**      the year of the record
- **name**      a common name for the storm; these are sometimes used for different storms in different years
- **letter**    the first letter of the name; storms are (usually) named in alphabetical order
- **doy**       the day of the year (1-365) of the record
- **hour**      the hour of the day (0-23) of the record in Eastern time 
- **lat**       latitude of the record in degrees
- **lon**       longitude of the record in degrees
- **status**    a two-digit status code of the storm system; see "storm_codes.csv"
- **category**  for hurricanes (status == "HU"), a number giving the category of the storm from 0-5
- **wind**      the observed sustained wind speed in miles per hour

There are also two metadata tables. One provides an automatically determined 
estimate of the gender of a storm's name. The last column gives a confidence
score.

```{r}
storm_gender <- read_csv("../data/storm_gender.csv")
storm_gender
```

There is a column in the main data set called `status` that described the type
of storm with a two letter code. The last table here provides a full name for
each of these codes:

```{r}
storm_codes <- read_csv("../data/storm_codes.csv")
storm_codes
```

If you have further questions about what these features mean, please let me know.

## Questions

### Question 1:

Produce a new table with one row for each storm in the data set that provides
the maximum wind speed the storm achieved over its lifetime.

```{r, question-01}

```

### Question 2: 

Hurricanes are labeled as being one of six different categories based on their
sustained wind speed. There is a column in the `storms` dataset called
`category` that provides this information. Create a new data set that has one
row for each hurricane category (status is equal to "HU") that shows the minimum
and maximum wind speed of all hurricanes in the data set with that category. 

```{r, question-02}

```

### Question 3: 

The data includes the day of the year (doy) and the hour of the day for each 
measurement. Most storms have records every six hours, but there is some 
variation here.

Create a new data set that has a unit of observation equal to the day of the 
year that counts the total number of storms since 1950 that were labelled as a
hurricane (status is equal to "HU") on midnight (hour is equal to 0) for a given
day. Note: You do **not** need to have rows in the dataeset representing days of
the year for  which there were never recorded hurricanes.

```{r, question-03}

```

### Question 4: 

Named Atlantic storms are given names in alphabetical order. The specific names
change from year to year. There is a column in the data that gives the letter
associated with each storm. Create a new data set that has two columns, letter 
and n, which indicate the number of years since 1950 for which the given letter
was the last letter used in that year. For example, if there were 9 years where
the last letter used was H, there should be a row with the letter H in one 
column and the number 9 in the other.

Note: my data has been filtered to exclude some storms, such as those with 
Greek letters, so do not expect these results to exactly match other sources.

```{r, question-04}

```

### Question 5:

In addition to the storms data, I also provided a small metadata table with
the predicted gender of each storm name (these are automatically generated
and may be incorrect). A few storms are not listed in the set of gender names.

I want you to create a table with two rows that shows the average maximum wind
speeds of the female storms in one row and male storms in the other row. Note:
By "average maximum", I mean that you should first compute each storm's maximum
wind speed and *then* take the average of these maximum values.

```{r, question-05}

```

### Question 6: 

I also provided a metadata table describing the meaning of each of the status
codes in the data (I already told you about "HU" being hurricane, but there are
several others). For this question, you need to produce a table that counts the
number of measurements with each status type with the full status names joined
into the output. Also, arrange the table from the most frequent to the least
frequent readings. Note: You must use a join function here for full credit.

```{r, question-06}

```

### Question 7: 

Produce a data set that is a subset of `storms` where we only have the
measurement where each storm first attains its largest wind speed. The columns
in the new data should be the same names as in the `storms` data. 

```{r, question-07}

```

### Question 8:

For this question, you need to produce a table with one row for each year in 
the `storms` data. There will be a column for each starting letter in the data
(note: the letters Q, U, X, Y, Z are not used). The values in the table should
give the maximum wind speed of the corresponding storm for each year.

As should be apparent from the question, this question requires a pivot. Add
the optional argument `names_sort = TRUE` to the pivot function to make the
columns easier to read. Your final table **should** have missing values.

For example, the 1950 storm named Baker (letter "B") had a maximum wind speed
of 90 mph. Therefore, the row corresponding to the year 1950 should have a 
value 90 in the column named "B".

```{r, question-08}

```

### Question 9:

As in the previous question, you need to produce a data set with one row for
each year in the storms data. Here, however, columns will align with the status
codes. The values in the table will give the number of records for each status
code in each year. This require another pivot. You do not need to add the
`names_sort = TRUE` option but should add the option `values_fill = 0`
to fill in the missing values with zeros.

```{r, question-09}

```

