---
title: "Exam 01: Takehome Portion (Solutions)"
output:
  html_document:
    theme: cosmo
    highlight: zenburn
    css: "../css/note-style.css"
---

## Instructions

Deadline: Monday, 12 February 2024 at the start of class

This exam is the same format as the notebooks we have been completing in class.
In order to complete it correctly, please keep in mind that:

1. In each section there is exactly one plot that you need to create. It will
often require using data verbs to create the plot, but it is up to you how 
to do that.

2. There are nine questions on this exam. Each is worth 10 points. An 
additional 10 points is assigned based on your code formatting across the
entire exam. This take-home exam will count for half of your Exam 01 grade.

3. You must Knit the file to an HTML format, print the file, and then bring
the exam to class on Wednesday. Some questions require you to build a plot 
with specific colors; you do not need to print the HTML in color. I will be
able to tell if you did the correct thing based on the code.

4. You may use any static resources, such as course notes and external
websites, but you may not discuss the exam with classmates anyone else.

5. I am happy to answer clarifying questions about the exam or to help you
with unexpected R errors. However, I will not answer content-based questions 
after the exam is posted. Note that I may not be able to answer questions 
sent after 8pm on Tuesday night. 

6. The exam should take no more than 2 hours, but you may use as much time 
as you need.

7. Personal computer issues is not an excuse for not completing the exam on
time. As a back up, you can use the computers in the Jepson computer lab or
in the library. 

Good luck!

## Weather in NYC

The data set for this exam consists of a single year of observed weather data
from New York City. Read in the data set with this command:

```{r}
source("../funs/funs.R")
weather <- read_csv("../data/weather-nyc.csv")
```

Here is a data dictionary for the features:

- **date**:                  the date of the weather record, formatted YYYY-MM-DD
- **actual_mean_temp**:      the measured average temperature for that day (Celsius)
- **actual_min_temp**:       the measured minimum temperature for that day (Celsius)
- **actual_max_temp**:       the measured maximum temperature for that day (Celsius)
- **average_min_temp**:      the average minimum temperature on that day since 1880 (Celsius)
- **average_max_temp**:      the average maximum temperature on that day since 1880 (Celsius)
- **record_min_temp**:       the lowest ever temperature on that day since 1880 (Celsius)
- **record_max_temp**:       the highest ever temperature on that day since 1880 (Celsius)
- **record_min_temp_year**:  the year that the lowest ever temperature occurred
- **record_max_temp_year**:  the year that the highest ever temperature occurred
- **actual_precipitation**:  the measured amount of rain or snow for that day (cm)
- **average_precipitation**: the average amount of rain or snow on that day since 1880 (cm)
- **record_precipitation**:  the highest amount of rain or snow on that day since 1880 (cm)
- **month**:                 month of the year as a string
- **description**:           text description of the weather, either "dry", "rain", or "snow"

If you have questions about what these features mean, please let me know.

## 1. A straightforward line plot

In the first plot, create a line plot (i.e., use `geom_line`) with date
on the x-axis and the ACTUAL maximum temperature on the y-axis. 

```{r, question-01}

```

## 2. Line plot with fixed color

Now, create a line plot (i.e., use `geom_line`) with date on the x-axis and the
AVERAGE maximum temperature on the y-axis. Make the line the color "red".

```{r, question-02}

```

## 3. Two Lines

Create a plot that layers your solutions to the first two questions into a 
single plot. In other words, it will have date on the x-axis and two lines: one
in black showing the actual maximum temperature and one in red showing the 
average maximum temperature. 

```{r, question-03}

```

## 4. Two lines on one plot with colors

In this plot, layer two line geometries (in other words, there will two lines)
that show the actual minimum temperature and the actual maximum temperature,
with date on the x-axis. Color the minimum temperature with the color named
"#5be5e5" (it's a pretty shade of blue) and the maximum temperature with
the color "#ff6666" (a pretty shade of red). 

```{r, question-04}

```

## 5. Variable Color

Now, construct a scatter plot with average maximum temperature on the x-axis 
and record maximum temperature on the y-axis. Color the points according to
the month of the year using the color-blind friendly viridis scale.

Note that the ordering of the months will be alphabetical; this is not ideal
in general, but fine for now!

```{r, question-05}

```

## 6. Snow: Text Labels on a Subset

Create a scatter plot with with average maximum temperature on the x-axis 
and record maximum temperature on the y-axis. Color all the points with the
color "grey". Add a layer on top of this consisting of a scatter plot with only
those days for which it was snowing colored in "blue".

```{r, question-06}

```

## 7. Summer: Layer Call Out

Create a scatter plot with with average maximum temperature on the x-axis 
and record maximum temperature on the y-axis. Color all the points with the
color "grey". Add another layer consisting of a scatter plot with only those
days from the months "June", "July", and "August" using the color "#ff6666".

```{r, question-07}

```

## 8. Segments

In this plot you are going to make use of the geometry layer `geom_segment`. I 
showed this in the slides, but you may not have used it yourself. It creates a
set of short line segments and requires four aesthetics: the `x` and `y` aesthetics
control where the segment begins and the `xend` and `yend` aesthetics describe
what features determine where the segment ends. 

The plot should have date again on the x-axis. Using a geom_segment layer, 
create segments that start with a y-value of actual_max_temp and proceed down
to the y-value of actual_min_temp. That is, each day will have a vertical line
going from the high temperature to the low temperature. 

```{r, question-08}

```

## 9. Coldest Days

Create a line plot with date on the x-axis and actual maximum temperature on
the y-axis. Add a point layer showing the coldest 15 days in the data. Color
these points in the color "#5be5e5".

```{r, question-09}

```
