---
title: "Notebook 07"
output:
  html_document:
    theme: cosmo
    highlight: zenburn
    css: "../css/note-style.css"
---

## Getting Started

Before running this notebook, select "Session > Restart R and Clear Output" in
the menu above to start a new R session. This will clear any old data sets and
give us a blank slate to start with.

After starting a new session, run the following code chunk to load the
libraries and data that we will be working with today.

```{r, include=FALSE, message=FALSE}
source("../funs/funs.R")
```

I have set the options `include=FALSE` and `message=FALSE` to avoid cluttering
the solutions with all the output from this code.

## Practice

### Movies data

For this notebook, we will be creating a datasets, reading it into R, and
producing a couple of different data visualizations.

We will create a dataset as a class describing a set of movies. Each student
will be tasked with creating three rows of data. We will record the following
items in the GoogleSheets file linked to on the class website:

- student number
- title
- year of release
- budget
- box-office, total amount of money
- genre
- personal rating of the film, from 1-5 stars

Once you are finished, save the data set as a csv file, upload it to the "data"
directory in RStudio, change its name, and read it into R in the following block
of code:

```{r, message=FALSE}
movies <- read_csv("../data/movies.csv")
movies
```

Then, produce a scatterplot below with the budget on the x-axis, box-office
numbers on the y-axis, points colored by the year of release, and the size of
the points corresponding to your personal rating. Use a text repel layer to
label the movies by their title. 

```{r, warning=FALSE, question-01}

```

If you think it is needed, also include logarithmic scales for the x- and
y-axes.

### Labels and Themes

For the last few questions, let's finish a few loose ends about how to make
data visualizations in R. The function `labs()` is an important element that
can be added to any ggplot graphic. Inside the `labs()` function, you can
set aesthetics from the plot equal to a string to change the corresponding
label in the plot. For example, adding `labs(x = "HELLO!")` will change the
label of the x-axis to the string "HELLO!". In the code below, take the
previous code from above and add an appropriate label to the x-axis.

```{r, warning=FALSE, question-02}

```

Now, copy the code above and in the next block add appropriate labels for
the y-axis, the color aesthetic, and the size aesthetic.

```{r, warning=FALSE, question-03}

```

There are three special additional elements that we can set inside of 
`labs()` called `title` (the main title of the plot), `subtitle`,
and `caption` (the description at the bottom of the plot). Take the
code above and add a title, subtitle, and caption.

```{r, warning=FALSE, question-04}

```

As a final step, we can add another element called a theme to change
the additional visual elements of the plot, such as what fonts are 
used in the labels, and what colors are used in the interior parts
of the graphic. The default that I have set for you in class is
called `theme_minimal()`. It's great for a first-pass of some graphics.
In the plot below, copy your graphic from above and add one final
element called `theme_tufte()`.

```{r, warning=FALSE, question-05}

```

I like this theme a lot for creating graphics for publication because 
it is very clean and avoids using extra ink that I can make better use
of with manual annotations. There are more other themes and theme
options that you can read about the R documentation to change the way
a ggplot object looks. Several other R packages also provide additional
themes ready to use out of the box. 
