---
title: "Notebook 05"
output:
  html_document:
    theme: cosmo
    highlight: zenburn
    css: "../css/note-style.css"
---

## Getting Started

Before running this notebook, select "Session > Restart R and Clear Output" in
the menu above to start a new R session. This will clear any old data sets and
give us a blank slate to start with.

After starting a new session, run the following code chunk to load the
libraries and data that we will be working with today.

```{r, include=FALSE, message=FALSE}
source("../funs/funs.R")
```

I have set the options `include=FALSE` and `message=FALSE` to avoid cluttering
the solutions with all the output from this code.

# Practice

We will work with the largest cities data sets:

```{r, message=FALSE}
cities <- read_csv("../data/largest_cities.csv")
cities
```

We will now also work with the entire U.S. cities data set:

```{r, message=FALSE}
us <- read_csv("../data/us_city_population.csv") |>
  filter(!is.na(population))
us
```

Please refer to the previous notebooks for more information about these data
sets and how these data sets are organized.

## Summary Statistics

In the code block below, using the `summarize` function to compute the mean
population (`population`) in the `cities` dataset.

```{r, question-01}

```

## Grouped Summaries

Let's now try to use grouped summarize functions. There is a feature in the
`cities` data set called `city_definition`. It describes the kind of
administrative structure given to each city. Using a grouped summary, in the
code below tabulate how many times each city definition is used in the dataset.
Arrange the data in decreasing order from the most common to least common
definition.

```{r, question-02}

```

What city type is the most common in the dataset?
**Answer**:

Now, count and arrange in descending order the feature `koppen_code`.

```{r, question-03}

```

What Köppen Code is the most common in the dataset?
**Answer**:

Now, in the code below group by continent and paste together the
city names (`name`).

```{r, question-04}

```

You will probably have to scroll over to see the results.

Finally, in the code below group by continent, count the number of
cities in each continent, and pass this to a plot with a `geom_col` layer
to visualize the number of cities on each continent.

```{r, question-05}

```

## Summarize Trends in U.S. Cities Data

We will now turn to the U.S. cities dataset to perform some more involved uses
of the summary function. To start, group by the year feature and summarize the
dataset by taking the sum of the population in every city for each year (with
`sum`). Draw a plot with `geom_line` and `geom_point` to show the population
trend in these 300 U.S. cities over time.

```{r, question-06}

```

The population feature in this data set is given in thousands of people. In
2000 there were approximately 300 million people living in the Unite States.
Roughly what fraction of people in the year 2000 appear to have lived in one of
the largest 300 cities according to this plot? **Answer**: It looks like about
80 million live in these cities, so about 25%.

## Grouped Arrange and Slice

In the notes we used the `group_by` function to manipulate the summarize
function. However, the functions `arrange`, `slice`, and `filter` also respect
the grouping of a data set. This can be quite useful. For example, consider
grouping the US cities data set by year, arrange in descending order by
population, and then using `slice` to take the first five rows. This would
result in a data set that gives the five largest cities for each year in our
data set. Write the code to do this below and visually verify that it seems to
pick out five cities for each year:

```{r, question-07}

```

Starting with the code in the previous block, summarize the data set
by pasting together the city names.

```{r, question-08}

```

In the code below, write the code to select one row for each city corresponding
to the year that the city had its largest population. (Note: think about this
carefully before you start writing the code).

```{r, question-09}

```

It would be helpful to sort the dataset you created in the previous code block
by the year feature. That would let us see the cities that peaked earliest at
the top of the dataset. However, if we added an arrange function at the end of
the code you wrote nothing would happen because dataset still grouped by city.
We need to first ungroup the dataset with the `ungroup()` function. In the code
below, starting with what you wrote in the block above, ungroup the dataset and
arrange by year:

```{r, question-10}

```

You should see that four of the earliest cities to peak in population are in
Massachusetts. Each of these four cities are known for being industrial towns
will large mills. What are the names of these cities?
**Answer**:

Taking the code to create the dataset that you produced in the previous
question, produce a bar plot showing the number of cities with a peak
population in each decade.

```{r, question-11}

```

Try to identify the three different clusters of peak city sizes.
**Note**: There is a small cluster in 1920 corresponding the mill
cities in the Northeastern U.S., a cluster in the 1950s and 1960s
of cities that decreased in size due to white flight and gerrymandering,
and (the large cluster) of cities that are larger today that they have
ever been.

To finish, in the code below take the `us` dataset and use the filter
function to select only cities with a longitude greater than -125.
Select the row corresponding to each cities largest year of population
(as we did in one of the previous questions) and produce a
scatterplot with longitude on the x-axis and latitude on the y-axis.
Color the points according to the year that the city attained its largest
population and include a color-blind friendly color scale. Finally, sort the
data in descending order of year before plotting.

```{r, question-12}

```

Try to match up the clusters of years with locations on the map.
