---
title: "Notebook 01"
output:
  html_document:
    theme: simplex
    highlight: pygments
    css: "../css/note-style.css"
---

```{r, include=FALSE}
source("../funs/funs.R")
```

## Practice: Running R Code

In this notebook are some further practice questions to see how well you
understood the material from today.

In the code block below, make a variable named `fav_number` and set it
equal to your favorite number. (Note, the quotes are not part of the
variable name).

```{r, question-01}

```

In the code below, apply the function `log10` to your favorite
number. (Again, the quotes are not part of the function name).

```{r, question-02}

```

Note that the only point of the exercises here are to familiarize
you with running code in R, creating variables, and applying functions.
We will get to more interesting tasks soon!

## Practice: Running Functions

R contains several functions for producing (pseudo)-random variables. These are
useful, for example, when using R to run simulation models. For example the
function `runif` selects (by default) a set of random numbers between 0 and 1.
It takes one required argument named `n`, which indicates how many random
numbers should be generated. In the code below, use the `runif` function to
produce 100 random numbers. Verify that each time you run the code a different
set of numbers is produced:

```{r, question-03}

```

The `runif` function also has two optional parameters. These are named `min`
and `max`; they determine the lower and upper bounds from which random numbers
should be generated. By default these are set to 0 and 1. In the code below,
generate 100 random numbers between 50 and 100. Here, select the number of
random numbers *by position* (that is, without the code `n = 100`).

```{r, question-04}

```

A common concern for new R users when running the code above is the meaning
behind the numbers in square brackets to the left of the output. These are not
part of the output itself. Instead, they provide a counter telling you which
result number appears just to the left of right of it. The first line will
always start with `[1]` because it always starts with the first result. The
second line's number will depend on the width of your screen when you ran the
code.

## Practice: Using Pipes

We will make heavy use of the `|>`, called a pipe, operator this semester. It
is a great tool for data analysis because it allows us to apply a sequence of
functions to an input (usually a data set) without needing to create
intermediate results.

For absolutely no reason other than practice, let's assume that we want to take
your favorite number above and apply the base-10 logarithm, followed by the
sine (sin()) function, followed by the cosine (cos()) function. We will do
this three different ways. To start, apply these three function in a single
line of R code by nesting them inside one another:

```{r, question-05}

```

Next, do the same sequence of operations, with one function on each line using
a temporary variable named `x`:

```{r, question-06}

```

Finally, apply the three functions using a sequence of pipes.

```{r, question-07}

```

Look at the previous three blocks and code and consider which is easiest to
read and modify. Try to think about possible benefits of each approach.

## Practice: Largest Cities data set

We will make extensive use of a data set that has information about the largest
cities (by population) in the world. To read in and print out this data set, run
the following lines of code:

```{r, message=FALSE}
cities <- read_csv("../data/largest_cities.csv")
```

Looking at the data, try to answer the following questions.

How many rows are in the data set?
**Answer**:

What are the observations in the data set.
**Answer**:

List three of the features in the data set.
**Answer**:

The population features are given in millions of people. How many
people live in all of Tokyo?
**Answer**:

How people live in Atlanta (Note: you will have to open the data
explorer to see this data)?
**Answer**:

## Practice: Formatting

In the code block below, I wrote some code to add a new column to the data set
that describes the population density for each city (1000s of people per square
kilometer) and sorts from the most dense to the least dense. We will learn
this code over the next 6 chapters. For now, just run the code block and see
what happens:

```{r}
new_data<-mutate(cities,city_density=city_pop/city_area*1000)
new_data<-arrange(new_data,desc(city_density))
select(new_data,name,country,city_density)
```

Let's focus on formatting the code above correctly. Note I did not include any
spaces! Copy the code above in the block below and then add the correct spaces
into the code to make it match the style guide given in the notes.

```{r, question-08}

```

Make sure to run the code after you are done. What city in the data set has, on
average, the most people per square kilometer?
**Answer**:

Now, take the code above and reformat it to use pipes in place of the temporary
variable.

```{r, question-09}

```

Verify that the results are the same as the previous code.

## Practice: Food

I have also given you the food data set given in the notes. In the code below,
read this data set into R using the `read_csv` function.

```{r, question-10}

```

If you have some time left in class. Take a look at all of the data columns for
the food data and ask if you are unclear on what any of these features indicate.
