---
title: "Notebook 08"
output:
  html_document:
    theme: cosmo
    highlight: zenburn
    css: "../css/note-style.css"
---

## Getting Started

Before running this notebook, select "Session > Restart R and Clear Output" in
the menu above to start a new R session. This will clear any old data sets and
give us a blank slate to start with.

After starting a new session, run the following code chunk to load the
libraries that we will be working with today.

```{r, include=FALSE, message=FALSE}
source("../funs/funs.R")
```

I have set the options `include=FALSE` and `message=FALSE` to avoid cluttering
the solutions with all the output from this code.

## College Majors

The data for today is taken from the following paper by a senior economist
at the Federal Reserve Board:

> Webber, Douglas A. "Are college costs worth it? How ability, major, and
> debt affect the returns to schooling." *Economics of Education Review* 53
> (2016): 296-310.

He uses information from the 2014-2018 American Community Surveys (a large
dataset compiled by the U.S. Census Bureau) to estimate the lifetime earnings
of college graduate based on their undergraduate major. The dataset includes
not just a single number, but instead 99 data points for each major, which 
describes the distribution of estimated lifetime earnings for each major. 
You can read in the dataset with the following:

```{r, message=FALSE}
majors <- read_csv("../data/majors.csv")
majors
```

It contains three fields:

- **major**        the name of the major
- **percentile**   percentile of earnings; includes every value from 1-99
- **earnings**     estimated lifetime earnings in millions of dollars

I do not believe that the estimates account for inflation.

In this notebook we will start by creating several different plots that use
this data to try to understand the relationship between college major and
lifetime earnings. We will then take a step back and consider how we can apply
the Data Feminism concepts from the notes to consider the implications and
limits of our analysis.

### Density Plots

To start understanding this dataset, we are going to make use of a new type of
geometry called `geom_density`. It is a bit different than previous geometries
we have used because rather than setting the y-aesthetic, we let the geometry
calculate the y-values based on the distribution of the x-aesthetic values.

In the code block below, take on the data points from the major "Psychology"
and draw a ggplot graphic with `earnings` on the x-axis using a `geom_density()`
layer.

```{r, question-01}

```

I find density plots easier to read if we shade the area underneath them with
a color. In the code below adapt your first plot by setting the `fill` aesthetic
to the fixed value "black" and the `alpha` aesthetic to the fixed value `0.3`.

```{r, question-02}

```

Let's compare the distribution of earnings from Psychology majors to two other
majors. In the code below, take just the rows from the majors "Psychology",
"Music" and "Art History" and draw a plot with `earnings` on the x-axis using
a `geom_density` layer. Set the fill aesthetic to the feature `major` and the
alpha value to 0.3.

```{r, question-03}

```

Take note of how would you describe the differences and similarities between
the estimated lifetime earnings of these three majors?

### Plotting Percentiles

Let's now look at all of the majors in the data. In the plot below, take just 
the rows of the data corresponding to the 50th percentile of earnings. Draw a
scatterplot with earnings on the x-axis and major on the y-axis. Order the
majors from the smallest to the largest earnings.

```{r, fig.height=8, question-04}

```

The point of the data we have today is to go beyond a single number in order to
get a different picture of how major influences lifetime earnings. In the plot
below, create a new version of the previous plot, but include two dots for 
each major, one for the 25th percentile of earnings and another for the 75th
percentile of earnings. Color the 25th percentile in "olivedrab" and the 75th
percentile "navy". Order the majors based on the 75th percentile of earnings.

Note: There are at least two different ways to do this task. Either is fine,
but try to do it without creating a second dataset using `mutate` and `if_else`
to construct a new color variable.

```{r, fig.height=8, question-05}

```

Repeat the previous question but produce a plot showing the 10th and 90th
percentiles of each major.

```{r, fig.height=8, question-06}

```

The 10th percentile of earnings for Chemical Engineers is 2.63 million dollars
over their lifetime. In the plot below, draw a scatter plot that shows the 
percentage of people in each major that are expected to make at least 2.63
million dollars over the course of their life. Put major on the y-axis and
the percentage on the x-axis and order by the percentage.

```{r, fig.height=8, question-07}

```

### Comparing Median Proportions

In the final plot, let's compare the actual differences in the median earnings
of each major. Filter to only include the 50th percentile of earnings and 
create a new variable that measures the earnings of each major divided by the
median earnings of all the majors. Create a plot similar to the others above
where the majors are ordered from the lowest earnings to the highest earnings
with the proportion of median earnings on the x-axis. 

```{r, fig.height=8, question-08}

```

The shape won't be  much different than the previous plots, but the scale of
the x-axis is (to me) more interpretable than the abstract idea of "lifetime
earnings".

### Reflections

Data Feminism says that we should consider the values and power structures that
are baked into the data and data analyses that we do. What values seem to be 
expressed in the data here regarding college majors and life goals?
**Answer**:

Data Feminism also says that data is never neutral or non-political. The data
here was an attempt to add nuance to studies focused only on a single summary
showing the relationship between college majors and earnings. Can you think of
another quantitative dataset that would help augment the story told by the data
we looked at above?
**Answer**:

Can you think of a qualitative source of information that would help augment
the story told by the quantitative data?
**Answer**:

It should be clear from these plots that while there is a relationship between
college majors and lifetime earnings, the situation is a bit more complicated
than the story told with a single number (say, the median earnings by major).
Can you construct a story that would use some of the results here to encourage
someone to major in a field that interests them rather than overly optimizing
for major that have the highest typical salaries?
**Answer**:

It is also possible to turn the data visualizations here into a fairly negative
story about lifetime earnings. Try to describe in the most pessimistic way 
possible how you might summarize the data here. The goal of this exercise is to
show how data analysis is never neutral and the same results can be interpreted
very differently.
**Answer**:

Note that my solutions to the answers above are just some of the possible
responses that are possible and are certainly not exhaustive of all ideas and
reflections that you might have about the data.
