---
title: "16. Time Series Data and Window Functions"
output:
  html_document:
    theme: cosmo
    highlight: zenburn
    css: "../css/note-style.css"
---

```{r, include=FALSE, message=FALSE}
source("../funs/funs.R")
```

## French COVID-19

### Load the Data

We will be using the same data sets from the previous notebook:

```{r, message = FALSE}
dept <- read_sf(file.path("..", "data", "france_departement.geojson"))
pop <- read_csv(file.path("..", "data", "france_departement_population.csv"))
covid <- read_csv(file.path("..", "data", "france_departement_covid.csv"))
```

This time, we will look into the temporal components of the data
and see how they can be integrated into the spatial visualisations.

### Plotting Dates

Notice that the date variable in the `covid` dataset has a special data
type called "date". One some reflection, it seems reasonable to have a
seperate data type for dates because they act somewhat like a number
(ordered, some meaning of distances) and somewhat like a categorical
variable (fixed values). In this section we will explore how to work with
these data objects.

```{r}
covid
```

We have already worked a bit for date objects in plotting. When used as the
x-axis (or less commonly, y-axis), a date object will display okay with no
additional work on our part.

```{r}
covid %>%
  filter(departement == 69) %>%
  ggplot(aes(date, deceased)) +
    geom_line()
```

Often with dates, it is useful to manually modify the labels on the time-based
axis. We do this with `scale_x_date` (or its y-equivalent if time is on the
y-axis). Adding this scale to a plot on its own will have no effect, but we can
change the default by changing three parameters:

- `date_breaks` a string describing the frequency of the labels, such as
   "month" or "2 years"
- `date_minor_breaks` a string describing the frequency of the grid-lines
- `date_labels` format string: [strptime](https://rdrr.io/r/base/strptime.html)

You can set just a subset of these, depending on how much control you want over
the plot. For example, let's label the previous plot by showing a label for
every month, and using the pattern "%B" to show (just) the full month's name.

```{r}
covid %>%
  filter(departement == 69) %>%
  ggplot(aes(date, deceased)) +
    geom_line() +
    scale_x_date(date_breaks = "month", date_labels = "%B", date_minor_breaks = "month")
```

The R programming language is designed to be used in English, but often the
output of plots will be distributed in other languages. You will note in the
previous plot that the month names are in English. How do we change this? We
can change the *locale* used within R by using the `Sys.setlocale` function.
This data comes from France, so let's change our locale to French:

```{r}
Sys.setlocale(locale = "fr_FR.UTF-8") # macOS
# Sys.setlocale(locale = "French_France.1252") # Windows
```

Running the plot again shows the month names in French:

```{r}
covid %>%
  filter(departement == 69) %>%
  ggplot(aes(date, deceased)) +
    geom_line() +
    scale_x_date(date_breaks = "month", date_labels = "%B")
```

For the rest of the notebook, we will move back to English:

```{r}
Sys.setlocale(locale = "en_US.UTF-8")
```

You will note that I actually included this line the header of the notebook for
today in order to make sure that the notebook is running in an English/US locale.
On your personal machine this may be different.

### Working with date variables

There are a number of functions that allow us to manipulate date objects, most
frequently to extract a particular element of the date and save it as a new
variable within a call to `mutate`. For example, `year()`, `month()` and
`day()` return a numeric version of the corresponding date components:

```{r}
covid %>%
  mutate(year = year(date), month = month(date), day = day(date)) %>%
  select(date, year, month, day) %>%
  unique()
```

We can also do mathematical operations on dates; adding or subtracting
an integer changes the date by the specified amount of days:

```{r}
covid %>%
  mutate(date_next = date + 2) %>%
  select(date, date_next) %>%
  unique()
```

It can also be useful to create a temporary date object, for example to select
a subset of days. This is done with one of two functions, the `ymd` function,
which takes a single string formatted as "YYYY-MM-DD":

```{r}
covid %>%
  filter(between(date, ymd("2020-07-02"), ymd("2020-07-07"))) %>%
  select(date) %>%
  unique()
```

Or the `make_date` function, which takes seperate numeric arguments for the
year, month, and date. Note that the year, month, and day default to 1970,
1, and 1, respectively.

```{r}
covid %>%
  filter(between(date, make_date(2020, 07), make_date(2020, 07, 15))) %>%
  select(date) %>%
  unique()
```

The functions `make_date` and `ymd` can also be used within a mutate function
to create a date object when it is not automatically constructed.

Another useful function is `isoweek`, which returns a numeric description
of the week of the year.

```{r}
covid %>%
  mutate(week = isoweek(date)) %>%
  select(date, week) %>%
  unique()
```

And `wday` returns the day of the week. Setting the options `label = TRUE`
and `abbr = FALSE` returns these as an ordered factor:

```{r}
covid %>%
  mutate(wday = wday(date, label = TRUE, abbr = FALSE)) %>%
  select(date, wday) %>%
  unique()
```

Note that the same options (label and abbr) also exist for the `month()`
function. Also, we can set the locale of `wday` within the function itself
like this:

```{r}
covid %>%
  mutate(wday = wday(date, locale = "fr_FR.UTF-8", label = TRUE, abbr = FALSE)) %>%
  select(date, wday) %>%
  unique()
```

Otherwise, it uses the setting from `Sys.setlocale`. Note that setting the
option inside the function does not change the locale for subsequent plots or
functions.
