---
title: "18. Window Functions"
output:
  html_document:
    theme: cosmo
    highlight: zenburn
    css: "../css/note-style.css"
---

```{r, include=FALSE, message=FALSE}
source("../funs/funs.R")
```

## French COVID-19

### Load the Data

We will be using the same data sets from the previous notebooks:

```{r, message = FALSE}
dept <- read_sf(file.path("..", "data", "france_departement.geojson"))
pop <- read_csv(file.path("..", "data", "france_departement_population.csv"))
covid <- read_csv(file.path("..", "data", "france_departement_covid.csv"))
```

This time, we will look into the temporal components of the data
and see how they can be integrated into the spatial visualisations.

### Window Functions

We will now look at a completely different way to work with time data.
Rather than working directly with a date variable, we focus on working with an
ordered set of data and associating each row with the rows before and after it.
To start, let's grab just a single département from the data, make sure that
it is ordered by the date variable, and select just the variable hospitalized
and the date.

```{r}
covid_paris <- covid %>%
  filter(departement_name == "Paris") %>%
  arrange(date) %>%
  select(date, hosp = hospitalised) # shortened name to better display below
covid_paris
```

The function `lead` lets us associate a row in the data set with the
value of a variable in the *following* row of data:

```{r}
covid_paris %>%
  mutate(hosp_next = lead(hosp))
```

By default lead looks one row down, but we can set the option `n`
to select a different offset. For example, looking a week into
the future:

```{r}
covid_paris %>%
  mutate(hosp_next = lead(hosp), hosp_next_week = lead(hosp, n = 7))
```

The function `lag` works exactly the same way, but associates a row with
the value of a variable in a *preceding* row. Note that the first row of
data has a missing value because there is not previous row; we can change
that with the option `default` (the same thing happens at the end of the
data when using `lead`, it just was not as clear in the data print-out).

```{r}
covid_paris %>%
  mutate(hosp_last = lag(hosp), hosp_last_default = lag(hosp, default = 0))
```

Other functions exist to do more complex relationships between rows. One
that will be helpful for us here is `roll_meanr`, which takes the rolling
average of a variable for a fixed number of values back in the table.
For example, setting `n = 2` takes the average of each value of the
hospitalization variable from each row and the preceding row:

```{r}
covid_paris %>%
  mutate(h_mean = roll_meanr(hosp, n = 2))
```

There are variations `roll_meanl` and `roll_mean` for taking a rolling
average with the future and symmetrically on either side. The `r` option
(right) is the most appropriate for many kinds of time-series data.

We can see the effect of rolling averages in the following plot:

```{r, warning = FALSE}
covid_paris %>%
  mutate(
    h_mean_7 = roll_meanr(hosp, n = 7),
    h_sum_30 = roll_meanr(hosp, n = 30)
  ) %>%
  ggplot(aes(x = date, y = hosp)) +
    geom_line() +
    geom_line(aes(y = h_mean_7), linetype = "dashed") +
    geom_line(aes(y = h_sum_30), linetype = "dotted")
```

Notice that the peak occurs at a different date for the smoothed averages, but
they are less noisy than the raw counts.
