---
title: "17. Working with datetime Objects"
output:
  html_document:
    theme: cosmo
    highlight: zenburn
    css: "../css/note-style.css"
---

```{r, include=FALSE, message=FALSE}
source("../funs/funs.R")
```

## Chicago Data

### Load the Data

Let's load some data from the city of Chicago

```{r, message = FALSE}
comarea <- read_sf(file.path("..", "data", "chicago_community_areas.geojson"))
medical <- read_csv(file.path("..", "data", "chicago_medical_examiner_cases.csv"))
```

We will look into the temporal components of the data.

## Time and Datetime objects

### What is a datetime object?

The medical examiner data has two fields that describe specific times,
the time of the incident and the time of death. We call these *datetime*
objects because they describe a specific time on a specific day.

```{r}
medical
```

Notice that these are similar but a bit different than the date objects we had
in the previous notes. As with dates, there exist several functions to extract
particular components of datetime objects. These include all of the functions
for dates, as well as specific ones for time of day:

```{r}
medical %>%
  select(date_incident_iso) %>%
  mutate(
    year = year(date_incident_iso),
    hour = hour(date_incident_iso),
    minute = minute(date_incident_iso)
  )
```

Sometimes datetime data is given overly precisely, making it hard to
group and count the data. A helpful function to working with such
data is `floor_date`, which rounds down to the nearest interval:

```{r}
medical %>%
  select(date_death_iso) %>%
  mutate(hour_death_iso = floor_date(date_death_iso, "hour"))
```

You can use different intervals---such as "minute", "day", "week",
or "2 hour"---depending on your specific application.

### Creating datetime objects

As with date objects, it can be useful to create datetime values on
the fly. This can be done by specifying individual components with
`make_datetime` (defaults go to 1970-01-01 00:00):

```{r}
medical %>%
  arrange(date_incident_iso) %>%
  filter(date_incident_iso > make_datetime(2020, 3, 1, 5, 6))
```

Or specifying the entire value as a string:

```{r}
medical %>%
  arrange(date_incident_iso) %>%
  filter(date_incident_iso > ymd_hms("2020-03-01 05:06:00"))
```

These can be useful for manipulating existing dates or for filtering data.

### Time Differences

Adding (or subtracting) integers from a date object increases (or decreases)
the date by the given number of days. A datetime object works similarly, but
changes by a number of seconds:

```{r}
medical %>%
  select(date_incident_iso) %>%
  mutate(date_incident_iso_plus1 = date_incident_iso + 1)
```

We can also take the difference between two datetime objects; converting
the result to a number with `as.numeric` yields the number of seconds
between the time time periods. Here is the difference between the incident
datetime and the death datetime:

```{r}
medical %>%
  select(date_incident_iso, date_death_iso) %>%
  mutate(diff = as.numeric(date_death_iso - date_incident_iso))
```

### Time Objects

Finally, we also have the ability to create a *time* object. This records
a time without a corresponding date. It can be useful for when you want
arithmetic to behave properly (that is, wrap around at midnight) or if
you want to group/plot datetime objects using only their time component.

The function `as_hms` will create a time object from a date object:

```{r}
medical %>%
  select(date_death_iso) %>%
  mutate(time_death = as_hms(date_death_iso))
```

Finally, the function hms can be used to create a *time* object from
scratch.
