\documentclass[11pt, a4paper]{article}

\usepackage{fontspec}
\usepackage{geometry}
\usepackage{fancyhdr}
\usepackage[hidelinks]{hyperref}
\usepackage[normalem]{ulem}
\usepackage{multicol}

\geometry{
  top=3cm,
  bottom=3cm,
  left=3cm,
  right=3cm,
  marginparsep=4pt,
  marginparwidth=1cm
}

\renewcommand{\headrulewidth}{0pt}
\pagestyle{fancyplain}
\fancyhf{}
\lfoot{}
\rfoot{}

\setlength{\parindent}{0pt}
\setlength{\parskip}{0pt}

\usepackage{xunicode}
\defaultfontfeatures{Mapping=tex-text}

\begin{document}

\begin{center}
\textbf{DSST289: Introduction to Data Science --- Taylor Arnold --- Spring 2024}
\end{center}

\vspace{0.5cm}

\textbf{Website}: \texttt{https://statsmaths.github.io/dsst289-s24}

\bigskip

\textbf{Topics:}
Methods for collecting, manipulating, visualizing, exploring, and presenting
data.

\bigskip

\textbf{Format:}
The semester is broken into three, four-week units. The first unit focuses on
data visualization and the second unit focuses on data collection. In the final
unit we will explore applications to spatial and temporal datasets. The
last week of the semester is dedicated to a final project. Class meetings are
hands-on and interactive. Please bring a computer, pencil, and something to
write on to each class. Class materials can be found on the course website.

\bigskip

\textbf{Homework:}
All class meetings, other than exam days, have a reading posted on our website.
A few questions or activities are included at the end of each reading. These
should be completed before class, with self-reported completion recorded
through the preclass form found at the top of the course website. Please bring
the written responses to class. Homework is graded based on the percentage of
assignments that were succesfully submitted.

\bigskip

\textbf{Exams:}
There are three midterm exams. Each has a take-home open-book component and 
a closed-book in-class component given on Wednesday. The take-home will be
distributed in advance of the in-class exam. Answers to the
take-home component should be printed out and handed in at the same time as
the in-class exam. A list of topics for the in-class exam will be posted on
the course website.

\bigskip

\textbf{Final Project:}
A final project will be due during the last week of class. The project will
focus on finding or creating a new dataset and applying the techniques learned
throughout the class to the analysis of it. The project will take the form of
a digital poster session and a one-page reflection. Detailed intructions will
be posted following fall break.

\bigskip

\textbf{Attendance:}
While regular attendance is expected, I understand that personal issues 
occasionally arise. Also, please do not come to class if you are feeling ill.
Mark on each pre-class form whether you attended the \textit{prior} class.
In the result of a negative response, briefly explain why you were out of
class. Excessive absenses (roughly, 5 or more) without a clear rational
will result in a warning followed by a possible reduction in your final grade.
If you are unable to take the in-class exams on the given date, it will be
replaced with an oral exam.

\bigskip

\textbf{Getting Help:}
We will usually have a lot of time in class to answer questions about the course
material. If questions arise outside of class, please also feel free to send
these by email. I am of course happy to schedule an office hours meeting for
any extended questions or personal concerns. Because I know everyone has a
busy schedule, I offer office hours primarily by appointment. Just send an
email with your availability at least 1 day before you'd like to meet. However,
note that I generally avoid answering conceptual questions about the homework
questions before they are due; just do your best and we will discuss in class.

\bigskip

\textbf{Final Grades:}
The three exam, final project, and homework grades are averaged (i.e., each
contributes 20\%). A letter grade is assigned as follows:
             A (93--100), A- (90--92),
B+ (87--89), B (83--86),  B- (80--82),
C+ (77--79), C (73--76),  C- (70--72), and F (0--69).

\end{document}
