---
title: "Exam 03: Takehome Portion"
output:
  html_document:
    theme: cosmo
    highlight: zenburn
    css: "../css/note-style.css"
---


## Instructions

Deadline: Monday, 15 April 2024 at the start of class

This exam is the same format as the notebooks we have been completing in class.
In order to complete it correctly, please keep in mind that:

1. In each section there is exactly one plot that you need to create. It will
often require using data verbs to create the plot, but it is up to you how 
to do that.

2. There are nine questions on this exam. Each is worth 10 points. An 
additional 10 points is assigned based on your code formatting across the
entire exam. This take-home exam will count for half of your Exam 01 grade.

3. You must Knit the file to an HTML format, print the file, and then bring
the exam to class on Wednesday. Some questions require you to build a plot 
with specific colors; you do not need to print the HTML in color. I will be
able to tell if you did the correct thing based on the code.

4. You may use any static resources, such as course notes and external
websites, but you may not discuss the exam with classmates anyone else.

5. I am happy to answer clarifying questions about the exam or to help you
with unexpected R errors. However, I will not answer content-based questions 
after the exam is posted. Note that I may not be able to answer questions 
sent after 8pm on Tuesday night. 

6. The exam should take no more than 2 hours, but you may use as much time 
as you need.

7. Personal computer issues is not an excuse for not completing the exam on
time. As a back up, you can use the computers in the Jepson computer lab or
in the library. 

Good luck!

## Tropical Storms

The data for this exam consists of information about tropical storms in the 
Atlantic Ocean between 1950 and 2020. It is a slightly revised version of 
the data from Exam II. It gives one row each time a particular storm is measured
(usually every 6 hours). Note that you will need to use both the year and name
of the storm to uniquely describe it.

```{r}
source("../funs/funs.R")
storms <- read_csv(file.path("../data", "storms2.csv"))
storms
```

The variables are:

- **year**    the year of the record
- **month**   the month of the year
- **day**     the day of the year
- **hour**    the hour of the day (0-23) of the record in Eastern time 
- **name**    a common name for the storm; these are sometimes used for different storms in different years
- **letter**  the first letter of the name; storms are (usually) named in alphabetical order
- **doy**     the day of the year (1-365) of the record
- **lat**     latitude of the record in degrees
- **lon**     longitude of the record in degrees
- **status**  a two-digit status code of the storm system; see "storm_codes.csv"
- **category**  for hurricanes (status == "HU"), a number giving the category of the storm from 0-5
- **wind**    the observed sustained wind speed in miles per hour

There are also three metadata tables. One provides an automatically
determined  estimate of the gender of a storm's name. The last column
gives a confidence score.

```{r}
storm_gender <- read_csv(file.path("../data", "storm_gender.csv"))
storm_gender
```

There is a column in the main data set called `status` that described the type
of storm with a two letter code. The last table here provides a full name for
each of these codes:

```{r}
storm_codes <- read_csv(file.path("../data", "storm_codes.csv"))
storm_codes
```

Finally, we also have a geographic data set of the states from the southeastern
U.S.

```{r}
state_se <- read_sf(file.path("../data", "state_se.geojson"))
state_se
```


If you have further questions about what these features mean, please let me know.

## Questions

### Question 1: Make datetime object [temporal]

Take the `storms` data and create a new column called `time` that is a date
time object created by the year, month, day, and hour of the storm's
measurement. The times are all given in the "America/New_York" timezone and
you should set this timezone in the code below. You will need this variable
in the following questions, so overwrite the data  `storms` with the new 
version that contains the `time` column.

```{r, question-01}

```

### Question 2: Plot datetime objects [temporal]

Take the storms data and take just the hurricanes (status code "HU") from the
year 2017. Group by the time variable you created in question 1 and count the
number of readings for each unique time. Create a plot with the data in which
the x-axis gives the time and the y-axis gives the number of hurricane readings
in 2017 at that time. Set the `date_breaks` and `date_minor_breaks` to both be
"1 week" and the date labels to be "%b %d".

```{r, warning = FALSE, question-02}

```


### Question 3: Create and plot date data [temporal]

Filter the data to include only category 5 hurricanes. Group the data by year
and hurricane name and compute the maximum wind speed of each storm. Create a
plot that has year on x-axis and max-wind speed on the y-axis. Use both a points
layer and a `geom_text_repel` with the name of the storm. Make sure that the 
x-axis is a date object and set `date_breaks` to be '5 year' and the date labels
to "%Y".

```{r, question-03}

```

### Question 4: Create and plot date data [temporal]

Take the storms data, using all records, and group by year and name. Summarize
the data by taking the difference between the maximum and minimum times for 
readings of each storm. Arrange in descending order, with the longest lasting
storms at the top. You should see one outlier at the top of the list resulting
from a bad data point.

```{r, question-04}

```

### Question 5: Rolling mean [temporal]

Group the storms data by the day of the year (`doy`) and count the number of
storms on each day of the year. Create a smoothed version of the count using
`roll_meanr` with a window size of 15. Plot the data with doy on the x-axis and
the count on the y-axis. Draw a solid line for the smoothed data and a dotted
(linetype = "dotted") for the raw data.

```{r, warning = FALSE, question-05}

```

### Question 6: Window functions [temporal]

Filter the storms data to include only readings from the 1995 storm "Felix". 
Take the lag of the variable `wind` and create a new variable called `color`
that is equal to "red" if a row of the data has an increased wind speed from the
previous row and to "navy" if the wind speed is the same or less. Plot the data
with time on the x-axis and the wind speed on the y-axis using a `geom_point` 
layer. Color the points based on the color variable (using `scale_color_identity`
to get the colors correct). Set the scale of the x-axis so that there are date
breaks and date minor breaks every day. Set `date_labels` equal to "%d". 

```{r, warning = FALSE, question-06}

```

### Question 7: Spatial polygons [temporal]

Create a plot of the states in `state_se` in the coordinate system 5069. Yes,
this is an easy question.

```{r, question-07}

```


### Question 8: Spatial points [spatial]

Select the storms data from the 2004 storm "Jeanne". Create a spatial plot with
the measurements of the storm's wind speed overlaid over the state boundaries.
Color the points based on the status code and use the 5069 coordinate system.

```{r, question-08}

```

### Question 9: Spatial join [spatial]

Filter the storms data to include only hurricanes (status code "HU"). Perform a
spatial join with the states data, group by the state names, and summarize to
count the number of hurricane readings in each state. Sort from the state with
the most readings to the state with the fewest.

```{r, question-09}

```
