---
title: "10. Table Pivots"
output:
  html_document:
    theme: cosmo
    highlight: zenburn
    css: "../css/note-style.css"
---

```{r, include=FALSE, message=FALSE}
source("../funs/funs.R")

food <- read_csv(file.path("..", "data", "food.csv"))
food_prices <- read_csv(file.path("..", "data", "food_prices.csv"))
```

## Motivation (read this section twice; thrice wouldn't hurt either)

In this notebook we introduce another set of methods for manipulating data sets.
Table pivots, which are related by not identical to the spreadsheet concept of
a *pivot table*, are a way or rearranging the values in a table without adding
or lossing any additional information. This is achieved by either making the
table longer (more rows, fewer columns) or wider (more columns, fewer rows).

What sort of situations would require going between two different formats with
a different number of rows? As an example, consider a hypothetical experiment
where we measure the heights of 100 plants every morning for two weeks. There
are two different ways to store this data. We could have 100 rows, one for each
plant, with features `height_day1`, `height_day2`, and so on all the way
through `height_day14`. Alternatively, we could have 1400 rows with just three
columns: an id for the plant, a feature for the day, and a feature for height.
Notice that both of these options capture the same information, but each
privileges a particular kind of analysis.

In the wider table format, it is straightforward to compute the amount that each
plot grew over the two weeks using a single mutate function. In the longer table
format, it would be straightforward to filter by a specific plant id and draw a
line plot showing the growth of a specific plant over the two week period. Both
drawing a plot with the wider table or computing the growth with the longer
table are possible, but require a surprising amount of work and code.

In this notebook we will introduce two new functions for alternating between
wider and longer formats for a data set. These are principles that will be
fundamental to several applications, particularly with text and temporal data
sets.

## Pivot wider

For these notes, we will make use of the `food_prices` data set. As shown in our
motivation example, pivoting is often a useful operation to apply when analyzing
data collected over time. The data set is organized with year as the observation
and each food type as a column.

```{r}
food_prices
```

This format makes it straightforward to compute the correlation between the
prices of different kinds of food items. A longer format for the data set would,
instead, have one row for each combination of year and food time.

In order to make this table longer, we will apply the `pivot_longer` function.
This function requires knowing which current features in the data set should
be turned into values in the output data set. Often, it is easier to describe
the set of values that will *not* be turned into values. Here, we indicate
(with a minus sign) that the year value should remain as a feature in the
output data set:

```{r}
food_prices |>
  pivot_longer(-c(year))
```

Already this looks close to what a long form of the food prices data set should
look like. One improvement that we can make is to set better column names,
which can be done by setting the options `names_to` and `values_to` in the
function call:

```{r}
food_prices |>
  pivot_longer(-c(year), names_to = "food", values_to = "price")
```

The longer form of the data set makes it much easier to do some kinds of
analysis. For example, we can draw a line chart of all of the food prices with
a single graphics layer:

```{r}
food_prices |>
  pivot_longer(-year, names_to = "food", values_to = "price") |>
  ggplot() +
    geom_line(aes(x = year, y = price, color = food))
```

Drawing this plot with the original data set would require manually including a
layer for each food type, selecting their colors, and building a manual legend.
The alternative using the longer table is certainly the preferred approach.

## Pivot wider

To illustrate making a table wider, let's create a new data set consisting of
the long format of the food prices data set from just the years 1950 and 1975:

```{r}
food_prices_long <- food_prices |>
  pivot_longer(-year, names_to = "food", values_to = "price") |>
  filter(year %in% c(1950, 1975))
```

As described in our motivating example, it makes sense for some analyses to make
each time point a column in a wider data set. To do this, we use the
`pivot_wider` function. We need to indicate the current feature contains the
values that will become new columns and the feature from which to take the
values for the new column from. Here, the names will come from the `years`
column (we want a new column for 1950 and another one for 1975) and the values
will be filled in with prices.

```{r}
food_prices_long |>
  pivot_wider(names_from = year, values_from = price)
```

One problem with the default output is that the column names now start with a
number, which is not an allowed variable name in R. This makes it awkward to
work with the data set; it is better to add a prefix to the names to make them
valid. This can be done by setting the `names_prefix` option in the
`pivot_wider` function.

```{r}
food_prices_long |>
  pivot_wider(
    names_from = year, values_from = price, names_prefix = "year_"
  )
```

This new form of the data set makes it straightforward to plot the price of each
food type in 1975 as a function of its price in 1950.

```{r}
food_prices_long |>
  pivot_wider(names_from = year, values_from = price, names_prefix = "year_") |>
  ggplot() +
    geom_point(aes(x = year_1950, y = year_1975)) +
    geom_text_repel(aes(x = year_1950, y = year_1975, label = food))
```

We can add some of the polishing touches mentioned in Notebook07 to make the
plot even more readable.

```{r}
food_prices_long |>
  pivot_wider(names_from = year, values_from = price, names_prefix = "year_") |>
  mutate(food = stri_trans_totitle(food)) |>
  ggplot() +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey") +
    geom_point(aes(x = year_1950, y = year_1975)) +
    geom_text_repel(aes(x = year_1950, y = year_1975, label = food)) +
    labs(x = "Price Index (1950)", y = "Price Index (1975)")
```

In this new plot, can you see what products got much more expensive, much less
expensive, and stayed about the same from 1950 to 1975?

## Patterns for table pivots

The syntax for making tables wider or longer is, on the surface, not much more
complex than other table verbs that we have covered in this text. The biggest
challenges with table pivots are identifying when they will simplify an analysis
and not over-using them. The best way to avoid these issues is to store your
data in the longest format that makes sense for your data. For example, in the
motivating example about plant growth, it is better to store the data with 1400
rows and 3 columns.

Storing data in a longer format has a number of benefits. Reducing the number of
columns makes it easier to document the (smaller set of) features with a
well-written data dictionary. Also, while avoided in our simple examples within
this notebook, pivoting wider also often requires less code and results in fewer
bugs. Several of these are illustrated in the practice exercises.

Perhaps the biggest benefit of storing data in a longer format is to avoid the
potentially complex chain of operations required to make the plot at the end of
the previous section. The original data set is stored with years as rows and
items as columns. Producing the plot requires thinking of years and columns and
items as rows; this needed us to first pivot longer and then pivot wider.
Keeping data in a longer format avoids the need for double pivots, while also
making the different kinds of analysis (item and year, year by item, item by
  year) all reasonable accessible.

In my experience, some social scientists have a bad habit of storing data in
extremely wide and unwieldy formats. It seems to be something that comes from
the way that feels natural to organize survey and time-series data sets. For
example, the U.S. Census Bureau produces data sets that often have hundreds of
columns. Keep this in mind as you collect your own data, or work with external
sources. Often an analysis that looks difficult at first will appear quite
straightforward after pivoting your data set to a longer format.

## Homework Questions

Consider the following subset of the `hans` dataset:

```{r, echo = FALSE, message = FALSE}
hans <- read_csv("../data/hans_roslin.csv") 
hans <- hans |>
  filter(year %in% c(1957, 2007)) |>
  filter(country %in% c("United States", "France", "China", "Mexico")) |>
  select(country, year, gdp)
hans
```

Answer the following four questions:

1. Re-write the table that would result from pivoting the table wider by
turning the countries into columns.
2. Re-write the table that would result from pivoting the table wider by
turning the years into columns.
3. Sketch a plot showing a graphic that would be easy to make with the data
from question one, but hard from the original data.
4. Sketch a plot showing a graphic that would be easy to make with the data
from question two, but hard from the original data.

## Homework Answers

1. **Re-write the table that would result from pivoting the table wider by
turning the countries into columns.**

```{r, echo = FALSE}
hans |>
  pivot_wider(names_from = country, values_from = gdp)
```

2. **Re-write the table that would result from pivoting the table wider by
turning the years into columns.**

```{r, echo = FALSE}
hans |>
  pivot_wider(names_from = year, values_from = gdp, names_prefix = "year_")
```

3. **Sketch a plot showing a graphic that would be easy to make with the data
from question one, but hard from the original data.**

```{r, echo = FALSE}
hans |>
  pivot_wider(names_from = country, values_from = gdp) |>
  ggplot(aes(China, France)) +
    geom_point() +
    geom_text_repel(aes(label = year)) +
    labs(x = "GDP of China", y = "GDP of France")
```

4. **Sketch a plot showing a graphic that would be easy to make with the data
from question two, but hard from the original data.**

```{r, echo = FALSE}
hans |>
  pivot_wider(names_from = year, values_from = gdp, names_prefix = "year_") |>
  ggplot(aes(year_1957, year_2007)) +
    geom_point() +
    geom_text_repel(aes(label = country)) +
    labs(x = "GDP in 1957", y = "GDP in 2007")
```