---
title: "09. Table Joins"
output:
  html_document:
    theme: cosmo
    highlight: zenburn
    css: "../css/note-style.css"
---

```{r, include=FALSE, message=FALSE}
source("../funs/funs.R")

food <- read_csv(file.path("..", "data", "food.csv"))
food_prices <- read_csv(file.path("..", "data", "food_prices.csv"))
```

## Overview (read this section twice)

These notes all focus on the core concept of a *primary key*. A primary key
is a feature or group of features (in other words, a column or group of 
columns) in a dataset that uniquely identifies a row of data.
In our food dataset, the feature `item` is a primary key. In the `us`
dataset, the `city` and `year` columns together form a primary key. We need
both to uniquely identify one observation.

Likewise, a *foreign key* is the appearance of a primary key within a
different dataset. The food group feature in the `food` dataset, for example,
could be a foreign key if we had another table with one row describing
information about each food group.

A primary key and the corresponding foreign key in another table form a
*relation*. Typically a relation maps a single row in one dataset to many rows
in another.

A table join is a way of combining two tables based on relations. The goal is
to match up a foreign key in one table with the primary key in another table,
to add new features from one dataset into another dataset. Several examples
will make this notion more clear.

## Joining data by key

As an example of performing table joins, we will start with a pared down
version of our foods data:

```{r, message=FALSE}
food_sml <- food |>
  select(item, food_group, calories)

food_sml
```

Now, consider the following new dataset corresponding to dietary restrictions
associated with different food groups:

```{r, message=FALSE}
diet <- read_csv(file.path("..", "data", "food_diet_restrictions.csv"))
diet
```

In the `diet` table, the food group feature is a primary key. Within the
`food` table, it is a foreign key. What we would like to do is to combine these
datasets by matching up rows that have the same values in the corresponding
columns.

For example, it could be helpful to include the columns `vegan`, `vegetarian`,
and `pescatarian` in the foods dataset. In order to do this, we need to
associate a row in the food dataset with the row in the diet dataset that
contains the same value of the feature `food_group`. We do this using the
function `left_join`; we pipe in the larger dataset that we want to join columns
to, provide the name of the table to grab columns from, and indicate what
column will be used to combine the two datasets.

```{r}
food_sml |>
  left_join(diet, by = c("food_group"))
```

And that is it! We now have the dietary restrictions added into the food
dataset. We can now compute models, summaries, and construct visualizations
based on the new metadata associated with each food group.

Now, let's consider a third dataset containing information about food recipes,
one for a Pot Roast and another for Guacamole:

```{r, message=FALSE}
recipes <- read_csv(file.path("..", "data", "food_recipes.csv"))
recipes
```

In this dataset, ingredient is a foreign key corresponding to the primary key
"item" in the foods dataset. For the recipes data, the primary key requires two
columns: both recipe and ingredient are needed to describe each row uniquely.

We can now try adding the calories information from the `food` dataset into the
recipes dataset. The complication here is that the column we want to join on has
a different name in each dataset ("ingredient" versus "item"). To specify this,
specify the two different names within the `by =` argument of the join function:

```{r}
recipes |>
  left_join(food_sml, by = c("ingredient" = "item"))
```

Now, we have a dataset that has added the food group and calories information
to our recipe dataset. Notice that the ingredient "Bay Leaf" is not in our food
dataset. Because of this, the food group and calories information for this
ingredient are filled in with missing values.

Notice that rows of data about food items in the dataset `food` that  are not
in any recipe are not included in the output. This is where the term "left"
comes from: all rows from the "left" (e.g., first) table are returned, but
non-matching rows in the second (e.g., "right") table are dropped.

## Inner and outer joins

There are four mutating joins available in **dplyr**. These have the
same syntax. If the join key is a primary key in one table and there
are no missing matches, they all function exactly the same. If the
join key is a primary key in one table and not all foreign keys in
the other table match an entry in the other, the following happens:

- `left_join` only non-matching rows in the first dataset are kept
- `right_join` only non-matching rows in the second dataset are kept
- `full_join` non-matching rows are included from either table
- `inner_join` only matching rows from both tables are included

For example, by replacing `left_join` with the function `inner_join`
we can modify the join from the last section to drop the missing term
(Bay Leaf) that is not contained in our food dataset:

```{r}
recipes |>
  inner_join(food_sml, by = c("ingredient" = "item"))
```

These four joins are called "mutating" joins because, like the mutate function,
they add columns to a dataset. There are also two "filter" joins that serve only
to select rows in the first (left) table. The function `semi_join` selects
rows in the first table that has a match in the second and `anti_join` selects
rows in the first table that do not have a match in the second. These are
useful for several data science tasks and will come up from time-to-time in our
applications.

Finally, there is another join called a "nested" join (`nest_join`). It has some
very interesting applications, but we avoid using it this semester.

## Homework Question

Hand write a dataset that consists of at least two different tables where
one table has a primary key that appears as foreign key on the second table.
The tables do not need to be large (10 rows between both tables is more than
enough) and the subject of the tables is up to you. Be creative and don't just
copy the example above.
