---
title: "05. Summarize Data"
output:
  html_document:
    theme: cosmo
    highlight: zenburn
    css: "../css/note-style.css"
---

```{r, include=FALSE, message=FALSE}
source("../funs/funs.R")

food <- read_csv(file.path("..", "data", "food.csv"))
food_prices <- read_csv(file.path("..", "data", "food_prices.csv"))
```

## The summarize verb

In the previous notebook we introduced the concept of data *verbs*. Four useful
examples were shown: `slice` and `filter` for taking a subset of rows,
`select` for taking a subset of columns, and `arrange` for reordering a
data set's rows. In this notebook we discuss another important verb,
`summarize` that collapses a data frame by using summary functions. Using this
verb is slightly more involved because we have to explain exactly how the data
should be summarized. We will introduce several helper functions to make this
process slightly easier.

Before describing the syntax for the summarize function, let's start with an
example. Here, we summarize our food data set by indicating the mean (average)
value of the sugar variable across the entire data set:

```{r}
food |>
  summarize(sugar_mean = mean(sugar))
```

Here we used the function `mean` inside of the function `summarize` to
produce the output. We specified which variable to compute the mean of by
giving its name inside of the `mean` function. Note that we need to define
what the name of the new variable is.

The results shows us that the average amount of sugar in a 100g portion of al
of the foods is 3.419g.

In order to compute multiple summaries at once, we can pass multiple functions
together are once. For example, here we compute the mean value of three
nutritional measurements:

```{r}
food |>
  summarize(
    sugar_mean = mean(sugar),
    calories_mean = mean(calories),
    vitamin_a_mean = mean(vitamin_a)
  )
```

Notice that R creates a new data set with the variable names we supplied above.
There are a number of other useful summary functions that work similarly,
such as `min`, `max`, `sum`, and `sd` (standard deviation).

## Grouped summaries

Summarizing the data set to a single row can be useful for understanding the
general trends in a data set or highlighting outliers. However, the real power
of the summary function comes when we pair it with grouped manipulations. This
will allow us to produce summaries *within* one or more grouping variables in
our data set.

When we use the `group_by` function, subsequent uses of the `summarize` function
will produce a summary that describes the properties of variables within the
variable used for grouping. The variable name(s) placed inside of the
`group_by` function indicate which variable(s) should be used for the groups.
For example, here we compute the mean number of calories of each food group:

```{r}
food |>
  group_by(food_group) |>
  summarize(calories_mean = mean(calories))
```

Notice that the output data set contains a column for the grouping variable
(`food_group`) and the summarized variable (`calories_mean`). The summarized
variable name is exactly the same as the non-grouped version and the final line
of code looks exactly the same as before. However, the output data set now
contains six rows, one for each food group.

Any summarization function that can be used for an ungrouped data set can also
be used for a grouped data set. Also, as before, we can put multiple summary
functions together to obtain different measurements of each group.

```{r}
food |>
  group_by(food_group) |>
  summarize(calories_mean = mean(calories), total_fat_mean = mean(total_fat))
```

Notice that the automatically produced variable names should make it clear
which column corresponds to each summary function.

## More summary functions

There are several additional summary functions that will be useful for
analyzing data. The function `n()` takes no arguments and returns a
valye that counts the total number of rows in the data set:

```{r}
food |>
  group_by(food_group) |>
  summarize(n = n())
```

The summary function `paste` collapses all of the values in a character
variable. For example, applying this summary it to the `item` category after
grouping by color, we can see all of the foods in the data set associated with
a specific color:

```{r}
food |>
  group_by(color) |>
  summarize(items = paste(item, collapse = "|"))
```

Do the foods correspond to the colors that you would expect?

## Geometries for summaries

We can use summarized data sets to produce new data visualizations. For
example, consider summarizing the average number of calories, average total fat,
and number of items in each food groups. We can take this data and construct a
scatter plot that shows the average fat and calories of each food group, along
with informative labels. Here's the code to make this visualization:

```{r}
food |>
  group_by(food_group) |>
  summarize(
    calories = mean(calories), total_fat = mean(total_fat), n = n()
  ) |>
  ggplot(aes(calories, total_fat)) +
    geom_point(aes(size = n), color = "grey85") +
    geom_text_repel(aes(label = food_group))
```

If this seems complex, don't worry! We are just putting together elements that
we have already covered, but it takes some practice before it becomes natural.

Scatterplots are often useful for displaying summarized information. There are
two additional `geom` types that often are useful specifically for the case of
summarized data sets.

If we want to create a bar plot, where the heights of the bars as given by a
column in the data set, we can use the `geom_col` layer type. For this, assign a
categorical variable to the `y`-aesthetic and the count variable to the
`x`-aesthetic (or vice-versa). For example, here is a bar plot showing the
number of items in each food group:

```{r}
food |>
  group_by(food_group) |>
  summarize(n = n()) |>
  ggplot(aes(n, food_group)) +
    geom_col()
```

There are two specific things to keep in mind with the `geom_col` layer. First,
there are two color-related `aes` categories: the border of the bars (`color`)
and the color used to shade the inside of the bars (`fill`). We can change
these exactly as we did with the single color value used with scatter plots.


```{r}
food |>
  group_by(food_group) |>
  summarize(n = n()) |>
  ggplot(aes(n, food_group)) +
    geom_col(color = "black", fill = "white")
```

I find that using a white fill color and a black border is often a good-looking
starting point. Also, you will notice that making the bars horizontal will make
it easier to read the category names when there are a larger number of
categories.

## Multiple groups

As mentioned above, it is possible to group a data set by multiple variables.
To do this, we can provide additional variables to the `group_by` function
separated by commas. For example, we could group the food data set into food
group and color, and summarize each combination of the two:

```{r}
food |>
  group_by(food_group, color) |>
  summarize(n = n(), calories = mean(calories))
```

Notice that now there is one row for each combination of the two groups.
However, there is no row for combinations that do not exist. So, there is no
row for pink dairy products nor for white fruit.


## Homework Questions

Let's now take all of the rows from the entire `hans` dataset:

```{r, echo = FALSE, message = FALSE}
hans <- read_csv("../data/hans_roslin.csv") 
hans |>
  select(country, continent, year, life_exp, gdp, pop)
```

For each of the following five questions, write the R code that would produce 
the desired data set:

1. Compute the average life expectancy from the year 2007.
2. Compute the average gdp from the year 2007.
3. Compute the average life expectancy of each continent in the year 2002.
4. Compute the total number of people living in each continent in the year 1957. Note that R has the function `sum()` that might be helpful here.
5. Compute the total number of countries in each continent.

We will talk about these questions together in class.
