---
title: "04. Organize Data"
output:
  html_document:
    theme: cosmo
    highlight: zenburn
    css: "../css/note-style.css"
---

```{r, include=FALSE, message=FALSE}
source("../funs/funs.R")

food <- read_csv(file.path("..", "data", "food.csv"))
food_prices <- read_csv(file.path("..", "data", "food_prices.csv"))
```

### Verbs

In these notes we are going to cover a set of functions that take
a data frame as an input and return a new version of the data frame. These
functions are called *verbs* and come from the **dplyr** package. If you are
familiar with running database queries, note that all of these verbs
map onto SQL commands. In fact, R can be set up so that
**dplyr** is called over a database rather than a local
data frame in memory.

There are over 40 verbs in the **dplyr** package, though most are
a minor variant or specific application of another verb. In this notebook
we will see only four of them:

- select a subset of rows from the original data set (`filter` and `slice`)
- select a subset of columns from the original data set (`select`)
- sort the rows of a data set (`arrange`)

In all verb functions, the first argument is the original data frame and the
output is a new data frame. Here, we will also see the functions `between` and
`%in%` to assist with the filtering command and `desc` to assist with arranging
the rows of a data set.

Note that **verbs do not modify the original data**; they operate on a copy of
the original data. We have to make an explicit name for the new data set if we
want to save it for use elsewhere.

## Choosing rows

It is often useful to take a subset of the rows of an existing data set, for
example if you want to build a model on a certain subpopulation or highlight a
particular part of the data in a plot. Perhaps the most straightforward way to
take a subset of rows is to indicate the specific row numbers that we want to
extract. In order to select rows by row numbers, we use the verb `slice`,
followed by the numbers of the rows we want separated by commas. Here is an
example taking the second, fifth, and seventh rows of the data:

```{r}
food |>
  slice(2, 5, 7)
```

As mentioned above, the code here does not change the data set `food` itself. It
still has all 61 rows of food contained in it. If we want to create a new
data set with just these three food item, we need to explicitly name and assign
it. For example, here is how we would create a data set of the first five food
items named `food_first_five`:

```{r}
food_first_five <- food |>
  slice(1, 2, 3, 4, 5)
```

There is a convenient a shorthand for selecting a range of row numbers, for
example every row from the tenth to the twentieth, by indicating the starting
and ending row number by a colon. Here, for example, is another way to select
the first five rows of the data set:

```{r}
food |>
  slice(1:5)
```

Another way to take a subset of our data is to select rows based on conditions
about the features in the data set. To do this we use the `filter` function,
which accepts a statement about features in the data set. Only rows where the
statements are true will be returned. For example, here is how we use the
`filter` command to select the foods that have more than 150 calories grams of
sugar in each serving:

```{r}
food |>
  filter(calories > 150)
```

The output data set has only 20 rows, compared to the 62 in the original data.
Other comparisons can be done with `<`, `>=` and `<=`. There is also a special
function called `between` that is often useful. For example, here are the rows
that have between 2 and 3 grams of total fat:

```{r}
food |>
  filter(between(total_fat, 2, 3))
```

If you want to filter on a categorical feature, you can use the `%in%`
operator to select specific categories. Here is the code to filter only the
fish and vegetable food groups:

```{r}
food |>
  filter(food_group %in% c("fish", "vegetable"))
```

As with the other verbs, we can chain together multiple calls to produce more
complex logic. For example, this code selects fruits that have more than 150
calories per serving:

```{r}
food |>
  filter(calories > 150) |>
  filter(food_group %in% c("fruit"))
```


Which results in a reduced data set with only 1 row (avocados). You can also
use `==` to test equality (`food_group == "fruit"`) or `!=` to test whether a
feature is *not* equal to a specific value.

It is also possible to create a chain of calls that then get piped into a call
to the `ggplot` function. For example, here is a plot of the fruits and
vegetables with the Avocado outlier removed (by limiting the maximum available
total fat).

```{r}
food |>
  filter(food_group %in% c("vegetable", "fruit")) |>
  filter(total_fat < 10) |>
  ggplot() +
    geom_point(aes(x = calories, y = total_fat, color = food_group)) +
    geom_text_repel(aes(x = calories, y = total_fat, label = item)) +
    scale_color_viridis_d()
```

The pattern of a starting with a data set, applying a number of
transformations, and then creating a visualization of the data
will become a common pattern in our analyses.

## Data and Layers

Now that we know how to create a subset of our data, let's use this new
knowledge to build some interesting data visualizations. To start, create a
data set that just consists of the food types that are in the meat food group:

```{r}
food_meat <- filter(food, food_group %in% c("meat"))
food_meat
```

One of the core ideas behind the Grammar of Graphics is that complex
visualizations can be constructed by layering relatively simply elements on top
of one another. What if we wanted to put together two layers where one layer
uses the `food` data set and the other uses `food_meat`? To do this, we can
*override* the default data set in a layer with the option `data =`. This will
use a different data set within a particular layer. For example, here is how we
can layer the meat data set on top of the rest of the food items.

```{r}
food |>
  ggplot() +
    geom_point(aes(x = calories, y = total_fat)) +
    geom_point(aes(x = calories, y = total_fat), data = food_meat)
```

This plot, however, does not look any different than it would if we were just
to plot all of the food together. The second layer of points just sits
unassumingly on top of the rest of the data. To rectify this, we can color each
layer a different color in order to distinguish them from one another. Let's
try to highlight the meat food group in a navy blue, while making the rest of
the points a light grey:

```{r}
food |>
  ggplot() +
    geom_point(aes(x = calories, y = total_fat), color = "grey85") +
    geom_point(aes(x = calories, y = total_fat), color = "navy", data = food_meat)
```

We now have a plot that shows exactly where the meats are relative to the other
food items. We can further build up the plot by showing the names of just these
rows of the dataset as well:

```{r}
food |>
  ggplot() +
    geom_point(aes(x = calories, y = total_fat), color = "grey85") +
    geom_point(aes(x = calories, y = total_fat), color = "navy", data = food_meat) +
    geom_text_repel(
      aes(x = calories, y = total_fat, label = item),
      color = "navy",
      data = food_meat
    )
```

Notice that the code is starting to get a bit more complicated and some of the
graphic layers are becoming a bit long. This is a good place to use the
shorthand notation to inherit aesthetics across layers, like this:

```{r}
food |>
  ggplot(aes(calories, total_fat)) +
    geom_point(color = "grey85") +
    geom_point(color = "navy", data = food_meat) +
    geom_text_repel(aes(label = item), color = "navy", data = food_meat)
```

Notice how a relatively small set of commands can be put together in different
ways to build a variety of plots. Already, we are making further progress
towards building informative and beautiful graphics in R!

## Selecting Columns

It is also possible to take a subset of the columns in a data set. To do this,
we make use of the verb `select`. We pass it the names of the features we want
to keep in the output data set, in the (possibly new) order that we want the
columns to be arranged in. Here, for example, is a new version of the foods
data set containing only the food item name followed by the amount of Vitamin A
and Vitamin C:

```{r}
food |>
  select(item, vitamin_a, vitamin_c)
```

We will not need to use the `select` verb as often as `filter` because for the
most part having extra features around does not effect data visualizations or
data models. However, it can be useful to displaying results. As we saw above,
the Vitamin A and Vitamin C columns were cut-off in the original output but are
not visible in the selected data set version. Removing and reordering unneeded
columns will also be useful in some of the advanced applications that are
discussed in the following classes.

## Arranging Rows

The verbs `slice` and `filter` determine a subset of rows to keep from the
original data set. The `arrange` verb, in contrast, keeps all of the original
data but re-orders its rows. Specifically, we give it one or more feature
names and it sorts the data by the first feature from smallest to largest
(or alphabetically for character feature). In the case of ties, the second
feature is used if given. More features can be given to further break
additional ties. Here is an example where we order the data set first by
`food_group` and then by `calories`:

```{r}
food |>
  arrange(food_group, calories)
```

In the new data set all of the dairy products come up first followed by the
fish products. Within each group, the items are sorted from the lowest to
highest number of calories.

The ordering can be reversed (i.e., from the highest to the lowest value) be
wrapping a feature name in the function `desc()`, such as this ordering
from the most saturated fat to the least:

```{r}
food |>
  arrange(desc(sat_fat))
```

In the result here, "Cheese" has been placed at the top of the data set,
followed by "Sour Cream" and "Lamb".

One particularly useful application of `arrange` is to pair it with the verb
`slice`. Here, for example, is the code to select the six foods in our data
set that have the highest amount of Vitamin A:

```{r}
food |>
  arrange(desc(vitamin_a)) |>
  slice(1:6)
```

By saving this data set, we could highlight these specific foods on top of a
plot comparing them to the remainder of the data.

## Homework Questions

Consider the following subset of the `hans` dataset we worked with in the
previous notebook:

```{r, echo = FALSE, message = FALSE}
hans <- read_csv("../data/hans_roslin.csv") 
hans |>
  filter(year == 2007) |>
  filter(country %in% c("United States", "France", "Canada", "Ireland",
                        "Netherlands", "Spain", "Portugal")) |>
  select(country, continent, life_exp, gdp)
```

For each of the following five questions, write the R code that would produce 
the desired data set:

1. Filter to include only those countries that have a life expectancy above 80 years.
2. Order the data set from the highest life expectancy to the lowest life expectancy.
3. Select only those rows that are located in Europe.
4. Create a data set that only has all seven rows of the original data but includes only the features `country` and `gdp`.
5. Create a data set that takes only the first two rows of the starting data.
