---
title: "Notebook 04 -- Solutions"
output:
  html_document:
    theme: cosmo
    highlight: zenburn
    css: "../css/note-style.css"
---

## Getting Started

Before running this notebook, select "Session > Restart R and Clear Output" in
the menu above to start a new R session. This will clear any old data sets and
give us a blank slate to start with.

After starting a new session, run the following code chunk to load the
libraries and data that we will be working with today.

```{r, include=FALSE, message=FALSE}
source("../funs/funs.R")
```

I have set the options `include=FALSE` and `message=FALSE` to avoid cluttering
the solutions with all the output from this code.

# Practice

We will work with the largest cities data sets:

```{r, message=FALSE}
cities <- read_csv("../data/largest_cities.csv")
cities
```

We will now also work with the entire U.S. cities data set:

```{r, message=FALSE}
us <- read_csv("../data/us_city_population.csv")
us
```

Please refer to the previous notebooks for more information about these data
sets and how these data sets are organized.

## Selecting rows and columns

Take the `cities` data set and select only the first 5 rows. (Note: for this
and other questions in the notebook, just write the code to produce the
required data. There is no need to save the results as a variable unless
specifically asked to)

```{r, question-01}
cities |>
  slice(1:5)
```

Now, filter the data set to include only those cities with a population
(`population`) over 10 million people.

```{r, question-02}
cities |>
  filter(population > 10)
```

How many cities have a population over 10 million people?
**Answer**: 33

In the code block below, extract the cities that have between 7 and
10 million people.

```{r, question-03}
cities |>
  filter(between(population, 7, 10))
```

How many cities have a population between 7 and 10 million
people?
**Answer**: 18

In the code block below, extract the cities in Asia that have between
7 and 10 million people

```{r, question-04}
cities |>
  filter(between(population, 7, 10)) |>
  filter(continent %in% c("Asia"))
```

How many Asian cities have a population between 7 and 10 million people?
**Answer**: 15

## Filtering plot

Take the data set `us`. Select only the year 1800 and remove rows that have a
population of zero.

```{r, question-05}
us |>
  filter(year == 1800) |>
  filter(population > 0)
```

How many cities have a non-zero population given for the year 1800?
**Answer**: 22

Now, starting with the data set you produced in the last chunk of code,
produce a plot with city on the x-axis and and population on the y-axis
using the `geom_col` layer:

```{r, question-06}
us |>
  filter(year == 1800) |>
  filter(population > 0) |>
  ggplot(aes(city, population)) +
    geom_col()
```

You will notice that the labels are too difficult to read because they run into
one another. Modify this plot in the code block below by flipping the x and y
aesthetics:

```{r, question-07}
us |>
  filter(year == 1800) |>
  filter(population > 0) |>
  ggplot(aes(population, city)) +
    geom_col()
```

How easy do you find this plot to read? What improvements might you want to
make to it? (Don't worry about what you can do right now; think about what would
make it easier to interpret).
**Answer**: I find it relatively easy to read, though it would be good to re-order the data in a more meaningful way, such as from the largest to the smallest city.

## Arranging rows

In the code below, extract the five European cities from `cities` that have the
largest population. (Note: this will take three different verbs).

```{r, question-08}
cities |>
  filter(continent == "Europe") |>
  arrange(desc(population)) |>
  slice(1:5)
```

Which European city in our data set has the largest population?
**Answer**: Istanbul (though, technically much of the city is actually in Asia).

## Selecting columns

Take the `cities` data set and filter only those cities in Africa and Europe.
Use the select function to see just the city name and the variable
`koppen_code`.

```{r, question-09}
cities |>
  filter(continent %in% c("Africa", "Europe")) |>
  select(name, koppen_code)
```

## Layering graphics

In the code below, create a data set called `cities_large_metro` by extracting
the rows of `cities` that have a metro region larger than the U.N. estimate of
the city's population.

```{r, question-10}
cities_large_metro <- cities |>
  filter(population < metro_pop)
```

Now, draw a scatter plot with longitude on the x-axis and latitude on the
y-axis. Show all of the rows in the data set using a light grey color. Then,
add points layer and a text layer (with the city names as labels) for the rows
in `cities_large_metro`. Pick a bold color for the points and labels.

```{r, question-11}
cities |>
  ggplot(aes(lon, lat)) +
    geom_point(color = "grey85") +
    geom_point(color = "olivedrab", data = cities_large_metro) +
    geom_text_repel(aes(label = name), color = "olivedrab", data = cities_large_metro)
```

Do you see any particular patterns about which cities have metropolitan regions
that are larger than the city's official U.N.-listed population? (Note: There is
at least one thing that becomes apparent, but you will probably not find
something that explains all of the data).
**Answer**: While there are cities in all of the major continents, there is a particular concentration of cities with large metro areas in the United States.

## A better bar plot

Let's now try to build a better bar plot. In the next code chunk, take the
`us` dataset, filter to just the year 1800 and cities with a non-zero
population, and arrange the cities by their population:

```{r, question-12}
us |>
  filter(year == 1800) |>
  filter(population > 0) |>
  arrange(population)
```

Now, starting with the code from the previous chunk, draw a plot using
`geom_col`, with population mapped to the x-axis and city mapped to the y-axis.
You will notice that this is no different from the previous plot you made above.

```{r, question-13}
us |>
  filter(year == 1800) |>
  filter(population > 0) |>
  arrange(population) |>
  ggplot(aes(population, city)) +
    geom_col()
```

Copy the code from above but wrap the variable `city` in the function
`fct_inorder` when setting it to the y aesthetic.

```{r, question-14}
us |>
  filter(year == 1800) |>
  filter(population > 0) |>
  arrange(population) |>
  ggplot(aes(population, fct_inorder(city))) +
    geom_col() +
    geom_text(aes(population + 3, fct_inorder(city), label = population), size = 3)
```
You should now notice that the cities are arranged by population. We will talk
more about this function in future notes.
