---
title: "Notebook 10 -- Solutions"
output:
  html_document:
    theme: cosmo
    highlight: zenburn
    css: "../css/note-style.css"
---

## Getting Started

Before running this notebook, select "Session > Restart R and Clear Output" in
the menu above to start a new R session. This will clear any old data sets and
give us a blank slate to start with.

After starting a new session, run the following code chunk to load the
libraries and data that we will be working with today.

```{r, include=FALSE, message=FALSE}
source("../funs/funs.R")
```

I have set the options `include=FALSE` and `message=FALSE` to avoid cluttering
the solutions with all the output from this code.

# Practice

## Load data sets

In this notebook we will work with the U.S. city population data set.

```{r, message=FALSE}
us <- read_csv("../data/us_city_population.csv")
```

Please refer to notebook02a for more information about the data set.

## Widening Data: Cities as Observations

The `us` cities data set is a perfect candidate for widening. In its default
form, there is one row for each unique combination of year and city. However,
some plots and analyses are easier if we create a data set with either time or
the cities.

Let's start by creating a widened data set where each row is a city. The
years will then be turned into columns. In the code below, create this
data set using the `pivot_wider` function, with names coming from the years
and values from the population. As we saw previously, use the names prefix
"year_" to avoid column names staring with a number.

```{r, question-01}
us |>
  pivot_wider(
    names_from = year, values_from = population, names_prefix = "year_"
   )
```

Below, take the data set created in the previous block and filter to include
only cities with a population above 600 thousand in 2010. (Note: Do this after
the pivot. It is much easier that way.) Then, plot the population of each city
in 1950 on the x-axis and the population in 2010 on the y-axis with points and
a text repel layer using the city names. Use logarithmic scales for the x- and
y-axes.

```{r, question-02}
us |>
  pivot_wider(
    names_from = year, values_from = population, names_prefix = "year_"
   ) |>
  filter(year_2010 > 600) |>
  ggplot(aes(year_1950, year_2010)) +
    geom_text_repel(aes(label = city)) +
    geom_point() +
    scale_x_log10() +
    scale_y_log10()
```

Another graph that we can create using the wide version of the data set is an
arrow plot showing the trend of each city over time. Below, piping the same
data in as above, create a plot using the `geom_segment` layer with city of
the y-axis and population on the x-axis. However, for this layer, you need to
describe two additional aesthetics: `xend` and `yend`. Set `x` to the
population in 1950 and `xend` to the population in 2010. The value for `yend`
should be equal to the city name, the same as `y`.

```{r, question-03}
us |>
  pivot_wider(
    names_from = year, values_from = population, names_prefix = "year_"
   ) |>
  filter(year_2010 > 600) |>
  ggplot(aes(year_1950, city)) +
    geom_segment(aes(xend = year_2010, yend = city)) +
    scale_x_log10()
```

One issue with this plot is that it does not show which direction the population
is changing (is it increasing or decreasing between 1950 and 2010?). We can fix
this by adding an option to the `geom_segment` layer that turns the segment
into an arrow. Namely:

   `arrow = arrow(length = unit(0.02, "npc"))`

Below, modify the previous plot to include an arrow. Note that this line goes
inside the `geom_segement` function, after the `aes` command. You may need to
indent into a new line as is sometimes done in the notes to make the code easily
readable. The `0.02` effects the size of the arrow and can be modified as you
see fit.

```{r, question-04}
us |>
  pivot_wider(
    names_from = year, values_from = population, names_prefix = "year_"
   ) |>
  filter(year_2010 > 600) |>
  ggplot(aes(year_1950, city)) +
    geom_segment(
      aes(xend = year_2010, yend = city),
      arrow = arrow(length = unit(0.02, "npc"))
    ) +
    scale_x_log10()
```

As one last step, let's add some color to the plot. Below, modify the code
from the previous chunk to include a mutate verb that sets a variable
called `color` to "#FF7F7F" if the population decreased between 1950 and
2010 and to "#ADD8E6" if the population increased. You should be able to
do this with a single call to `if_else`. Then, color the lines with this
color using `scale_color_identity`. Also, make the line size `1.5` to make
the colors stand out more. Finally, order the cities by their size in 1950.

```{r, question-05}
us |>
  pivot_wider(
    names_from = year, values_from = population, names_prefix = "year_"
   ) |>
  filter(year_2010 > 600) |>
  mutate(color = if_else(year_1950 > year_2010, "#FF7F7F", "#7FFF7F")) |>
  arrange(year_1950) |>
  mutate(city = fct_inorder(city)) |>
  ggplot(aes(year_1950, city)) +
    geom_segment(
      aes(xend = year_2010, yend = city, color = color),
      size = 1.5,
      arrow = arrow(length = unit(0.02, "npc"))
    ) +
    scale_x_log10() +
    scale_color_identity()
```

## Widening Data: Years as Observations

Another possibility for pivoting the `us` data set is to have the observations
be years and the variables be the cities In the code block below, try to
`pivot_wider` using names from the city variable and values from the
population. We do not need a variable prefix here.

```{r, question-06}
us |>
  pivot_wider(names_from = city, values_from = population)
```

You should see that there is a problem. The issue is that there are several
other variables tied to each specific city (lon, lat, and state) that make
the output data set have many missing values and far too many rows. Modify
the code above by first using the `select` function to grab only the three
variables `city`, `year`, and `population`.

```{r, question-07}
us |>
  select(city, year, population) |>
  pivot_wider(names_from = city, values_from = population)
```

You should now have a data set with 23 rows (one for each year) and 301
columns (a year column and 300 cities). The shape of this data set is
fine, but the column names are difficult to work with because they contain
spaces and commas. We can fix this by changing the city names using some
string processing. Run the following code to create a new version of the
us cities data called `us_cities`:

```{r}
us_cities <- us |>
  mutate(city = stri_trans_tolower(city)) |>
  mutate(city = stri_replace_all(city, "", regex = "[^a-z ]")) |>
  mutate(city = stri_replace_all(city, "_", fixed = " "))
```

Run the answer to your last question with the new dataset `us_cities` and
verify that the names are now "proper" R variable names:

```{r, question-08}
us_cities |>
  select(city, year, population) |>
  pivot_wider(names_from = city, values_from = population)
```

Starting with the code above, create a plot with one city (of your choosing) on
the x-axis, another city (of your choosing) on the y-axis using a text layer
where the label is given by the year. Use a `geom_path` layer and an text repel
layer labeled by year. This should allow you to see how the population of the
two cities change together over time.

```{r, question-09}
us_cities |>
  select(city, year, population) |>
  pivot_wider(names_from = city, values_from = population) |>
  ggplot(aes(boston_ma, richmond_va)) +
    geom_point(color = "grey85") +
    geom_path(color = "grey85") +
    geom_text_repel(aes(label = year))
```

Note that the `geom_path` geometry is different than the `geom_line` geometry.
The latter always connects dots from the smallest x-value to the largest
x-value, whereas the path geometry connects the dots in the order they appear
in the data.

## Closing Thoughts

The `us` data set is a nice example of how keeping data in a long format makes
it relatively easy to work with the data as-is or in various wider formats. In
this example, the cities as observations seems to allow more interesting plots.
However, the second format will be more useful for building the kinds of time
series models we introduce in the applications with temporal data sets.
