---
title: "Notebook 18 -- Solutions"
output:
  html_document:
    theme: cosmo
    highlight: zenburn
    css: "../css/note-style.css"
---

## Getting Started

Before running this notebook, select "Session > Restart R and Clear Output" in
the menu above to start a new R session. This will clear any old data sets and
give us a blank slate to start with.

After starting a new session, run the following code chunk to load the
libraries and data that we will be working with today.

```{r, include=FALSE, message=FALSE}
source("../funs/funs.R")
```

## Pandemic Data in France

Today, let's look at an application using data from the COVID-19 pandemic in
France. Today we will use several data sets. Load them in below and take a
moment to look at them.

```{r}
covid <- read_csv("../data/france_departement_covid.csv")
pop <- read_csv("../data/france_departement_population.csv")
cities <- read_csv("../data/france_cities.csv")
geo_fr <- read_sf("../data/france_departement.geojson") |>
  filter(departement <= "95")
```

As a starting point, produce a spatial plot of the `geo_fr` data with the 
default projection:

```{r, fig.height=6, question-01}
geo_fr |>
  ggplot() +
    geom_sf()
```

Next, modify your plot to use the coordinate system with CRS number 27561. Use
this transformation for all plots in the following questions using the French
dataset.

```{r, fig.height=6, question-02}
geo_fr |>
  st_transform(27561) |>
  ggplot() +
    geom_sf()
```

Now, filter the dataset `covid` to the date "2020-05-01" and left join it into
the `geo_fr` dataset by the key "departement". You should notice that the output
dataset no longer has the special spatial header attached to it.

```{r, question-03}
covid |>
  filter(date == "2020-05-01") |>
  left_join(geo_fr, by = "departement")
```

In the block below, copy your code from the previous question and add an extra
line with the function `st_as_sf()`. This will let R know that the resulting 
dataset should retain all of the spatial information.

```{r, question-04}
covid |>
  filter(date == "2020-05-01") |>
  left_join(geo_fr, by = "departement") |>
  st_as_sf() 
```

Now, starting with the code that you created above, plot a map of the French
Département according to the the number of people that were hospitalized one
1 May 2020. Use `scale_fill_distiller` to pick a good color scale.

```{r, fig.height=7, question-05}
covid |>
  filter(date == "2020-05-01") |>
  left_join(geo_fr, by = "departement") |>
  st_as_sf() |>
  st_transform(27561) |>
  ggplot() +
    geom_sf(aes(fill = hospitalised)) +
    scale_fill_distiller(
      trans = "log2", palette = "Spectral", guide = "legend", n.breaks = 10
    )
```

A problem with the above map is that it is not normalized by population. Adapt
the code above in the block below by joining in the `pop` dataset and instead 
plotting the hospitalization rate were 100k people.

```{r, fig.height=7, question-06}
covid |>
  filter(date == "2020-05-01") |>
  left_join(geo_fr, by = "departement") |>
  left_join(pop, by = "departement") |>
  mutate(rate = hospitalised / population * 100000) |>
  st_as_sf() |>
  st_transform(27561) |>
  ggplot() +
    geom_sf(aes(fill = rate)) +
    scale_fill_distiller(
      trans = "log2", palette = "Spectral", guide = "legend", n.breaks = 10
    )
```

We also have a data set of French cities called `cities`. Create a spatial
points version of the cities data and save it as an object called `cities_geo`.

```{r, question-07}
cities_geo <- cities |>
  st_as_sf(coords = c("lon", "lat"), crs = 4326, remove = FALSE)
```

Create a spatial plot with the `cities_geo` dataset that shows the labels of the
cities. You can use the options `size` and `nudge_y` to play around with the 
size and location of the city names. Don't forget the `coord_sf()` layer to
make the plot use a proper map coordinate system.

```{r, fig.height=7, warning = FALSE, question-08}
cities_geo |>
  st_transform(27561) |>
  ggplot() +
    geom_sf_text(aes(label = city), nudge_y = 0.1, size = 4) +
    coord_sf()
```

Now, put together the hopsitalization rate map with the cities data in a single
plot. Include both points and text labels for the cities. Spend a few minutes
trying to get the plot to look nice on your machine.

```{r, fig.height=7, warning = FALSE, question-09}
covid |>
  filter(date == "2020-05-01") |>
  left_join(geo_fr, by = "departement") |>
  left_join(pop, by = "departement") |>
  mutate(rate = hospitalised / population * 100000) |>
  st_as_sf() |>
  st_transform(27561) |>
  ggplot() +
    geom_sf(aes(fill = rate), alpha = 0.4, color = NA) +
    geom_sf(data = cities_geo) +
    geom_sf_text(aes(label = city), data = cities_geo, nudge_y = 20000, size = 2) +
    scale_fill_distiller(
      trans = "log2", palette = "Spectral", guide = "legend", n.breaks = 10
    )
```

## Spatial Lines: NYC Roads

Spatial data can also associate lines with each row rather than a single point
or an entire polygon. These are somewhat more rare, but interesting nonetheless.

Let's read in a dataset of roads in New York City. Take another moment to look
at the data print out and the data in the data viewer.

```{r}
nyc <- read_sf("../data/ny_roads.geojson")
nyc
```

Spatial lines can be plotted with `geom_sf` much like spatial points. As with
points, we can do this without specifying any x- or y-aesthetic. Try this here
with the `nyc` data; you should see a familiar shape of the city.

```{r, fig.height=8, question-10}
nyc |>
  ggplot() +
    geom_sf()
```

And again, as with points, we can change the color and size of the lines if 
needed. Below, repeat your plot of the NYC roads in a color of your choosing.

```{r, fig.height=8, question-11}
nyc |>
  ggplot() +
    geom_sf(color = "pink")
```

There are a number of functions that compute summaries of spatial data. For 
example if we type the following inside of a mutate verb:

  len = as.numeric(st_length(geometry))
  
We will get the length of a line geometry in meters. Try this below, sorting the
output from the longest to the shortest road in NYC:

```{r, fig.height=8, question-12}
nyc |>
  mutate(len = as.numeric(st_length(geometry))) |>
  arrange(desc(len))
```

Notice that some of the city roads have duplicates because there are multiple
segments of a road broken up by some other feature. 

Now, let's do something a bit more involved. In the block below, create a new
data set named `nyc_broad` that only contains segements of the street with a 
name equal to "Broadway".

```{r, question-13}
nyc_broad <- nyc |>
  filter(name == "Broadway")
```

Finally, create a map of the streets in NYC in which most streets are the color 
"grey85" but Broadway is in the color "orange". Also make Broadway have a size
of 2. Note: this can be done by layering two `geom_sf` layers.

```{r, fig.height=8, question-14}
nyc |>
  ggplot() +
    geom_sf(color = "grey85") +
    geom_sf(data = nyc_broad, color = "orange", size = 2)
```

